A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-15 tidyindex.qmd
rework on spatial processing and dimension reduction + proofread

| No. | Module                  | Input             | Operation                                                                                                                                          | Output                                                              |
|-------------|-------------|-------------|-----------------|-----------------|
| 1   | Temporal processing     | $x(t)$            | $f[x(.)]$                                                                                                                                          | $x^*(t^\prime)$ where $t\prime \in D_{t^\prime}$                    |
| 2   | Spatial processing      | $x(s)$            | aaaaa                                                                                                                                              | $x^\#(s^\prime)$                                                    |
| 3   | Variable transformation | $x(s; t)$         | $T(x(s;t))$                                                                                                                                        | $u(s;t)$ ????                                                       |
| 4   | Scaling                 | $x(s; t)$         | $[x(s;t) - \alpha]/\gamma$                                                                                                                         | $v(s;t)$ where bbbbb                                                |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $y \in \mathbb{R}^d$ and $d < p$            |
| 6   | Distribution fit        | $x(s; t)$         | $F_\mathbf{\theta}(x(s,t))$ where $F_{\mathbf{\theta}}(.)$ is the cumulated distribution function with distribution parameter(s) $\mathbf{\theta}$ | $p(s,t)$ where $p(.)$ denotes the probability value between 0 and 1 |
| 7   | Normalising             | $x(s; t)$         | $\Phi^{-1}[x(s, t)]$                                                                                                                               | $z(s; t)$ where $z \sim N(.,.)$                                     |
| 8   | Benchmarking            | $x(s; t)$         |                                                                                                                                                    | $b(s;t)$                                                            |
| 9   | Simplification          | $x(s; t)$         | ccccc                                                                                                                                              | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                              |

@@

`Computation`: The temporal processing module takes a single variable in a set of fixed locations, $x(t)$, as the input. This step *transforms* the original series into a new series to incorporate the temporal neighbouring effect: $x^*(.) = f[x(.)]$. In temporal processing, the temporal dimension $t$ may also transition to $t^\prime \subseteq D_{t^\prime}$, allowing for the decomposition of time into its constituent temporal components. For example, this can be used to break down daily data into monthly. This gives the output of the temporal processing module as $x^*(t^\prime)$.

An example of temporal processing is in Standardized Precipitation Index (SPI), where monthly precipitation is summed over a rolling window of size $k$, also known as the time scale. Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be written as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$. For SPI, the choice of the time scale $k$, is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).

@@

`Input`: $x^l_t(s)$

`Output`: $x^l_t(s^\prime)$ where $s^\prime \in D_{s^{\prime}}$

`Computation`: Spatial processing may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing. The process can be written as a general operation in the form of

```{=tex}
\begin{equation}
x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
\end{equation}
```
where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the associated parameters in the process and $d_{\theta}$ is the number of parameter of $\theta$. An example of spatial processing is to align variables collected in different resolutions. When variables are collected at different resolutions, analysts may choose to down-sample those in a finer resolution, $i$, to match those in a coarser resolution, $i^\prime$. This is a spatial aggregation and if aggregate using the mean, it can be written as

```{=tex}
\begin{equation}
g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}},
\end{equation}
```
where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid and $n_{i^\prime}$ is the number of observations falls into the coarser grid. Other examples of spatial processing include 1) borrowing information from neighbouring spatial locations to interpolate unobserved locations and 2) fusing variables from ground measures with satellite imageries.

@@

`Output`: $x^{l^\prime}(s; t)$, where $l^{\prime} = 1, 2, \cdots, L^{\prime}$

`Computation`: Dimension reduction summarises the multivariate information into univariate, which can be denoted as:

```{=tex}
\begin{equation}
x_p(\mathbf{s}; \mathbf{t}) \rightarrow x(\mathbf{s}; \mathbf{t}),
\end{equation}
```
where $p = 1, 2, \cdots, P$. The combination can be based on domain-specific knowledge, originated from theories describing the underlying physical process. For example, the SPEI uses a water balance model ($D = P - \text{PET}$) to calculate the difference series ($D$) from precipitation ($P$) and potential evapotranspiration ($\text{PET}$).

Another widely used approach is linear combination, which aggregates a collection of variables in a linear additive structure, expressed as: $$x(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),$$

where $\lambda_p$ denotes the weight assigned to variable $x_p$. Most indexes uses an equal weight that sum to 1, which allows each variable to contribute equally to the index. A linear combination can also be viewed as a linear projection of multivariate information by the weight vector. Projecting data from higher to lower dimension inevitably leads to information loss. For example, two countries can receive similar index, but their components could significantly differ -- one with average scores across all components and another with extreme scores the opposite ends. The selected set of weights need to be examined to understand its effect on the index and its implications for decision-making. To do this, analysts can vary the coefficients in the linear projection to observe how the index value and countries' ranking change. These changes can also be visualized using a method called tour by generating an animation of the projections among different sets of weights.

>>>>

| No. | Module                  | Input                                                       | Operation                                                                                                                                          | Output                                                                  |
|---------------|---------------|---------------|---------------|---------------|
| 1   | Temporal processing     | $x(t)$                                                      | $f_{\Psi}[x(.)]$                                                                                                                                   | $x^*(t^\prime)$ where $t\prime \in D_{t^\prime}$                        |
| 2   | Spatial processing      | $x(s)$                                                      | $g_{\Theta}[x(.)]$                                                                                                                                 | $x^\#(s^\prime)$ where $s^\prime \in D_{s^\prime}$                      |
| 3   | Variable transformation | $x(s; t)$                                                   | $T(x(s;t))$                                                                                                                                        | $u(s;t)$ ????                                                           |
| 4   | Scaling                 | $x(s; t)$                                                   | $[x(s;t) - \alpha]/\gamma$                                                                                                                         | $v(s;t)$ where bbbbb                                                    |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ where $\mathbf{x} \subseteq \mathbb{R}^p$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < p$ |
| 6   | Distribution fit        | $x(s; t)$                                                   | $F_\mathbf{\theta}(x(s,t))$ where $F_{\mathbf{\theta}}(.)$ is the cumulated distribution function with distribution parameter(s) $\mathbf{\theta}$ | $p(s,t)$ where $p(.)$ denotes the probability value between 0 and 1     |
| 7   | Normalising             | $x(s; t)$                                                   | $\Phi^{-1}[x(s, t)]$                                                                                                                               | $z(s; t)$ where $z \sim N(.,.)$                                         |
| 8   | Benchmarking            | $x(s; t)$                                                   |                                                                                                                                                    | $b(s;t)$                                                                |
| 9   | Simplification          | $x(s; t)$                                                   | ccccc                                                                                                                                              | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                  |

@@

`Computation`: The temporal processing module takes a single variable in a set of fixed locations, $x(t)$, as input. This step transforms the original series into a new series to incorporate the temporal neighbouring effect: $x^*(.) = f_{\Psi}[x(.)]$, where $\Psi$ is a vector of hyper-parameters associated with the transformation. In temporal processing, the transition of temporal dimension from $t$ to $t^\prime \subseteq D_{t^\prime}$ allows for the decomposition of time into its constituent temporal components. For example, to break down daily data into monthly.

In the Standardized Precipitation Index (SPI), temporal processing is used to sum the monthly precipitation over a rolling window of size $k$, also known as the time scale, prior to the distribution fitting and normalising module. Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated precipitation series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be expressed as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$. For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).

@@

`Input`: $x(s)$

`Output`: $x^\#(s^\prime)$ where $s^\prime \in D_{s^{\prime}}$

`Computation`: The spatial processing module takes a single variable with a fixed temporal dimension, $x(s)$, as input. This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^\#(.) = g_{\Theta}(x(.))$, where $\Theta$ is a vector of hyper-parameters associated with the transformation. The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions. This also includes the aggregation of variables into different levels, such as city, state, and countries scales.

<!-- may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing. The process can be written as a general operation in the form of -->

<!-- x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})), -->

<!-- where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the associated parameters in the process and $d_{\theta}$ is the number of parameter of $\theta$. An example of spatial processing is to align variables collected in different resolutions. When variables are collected at different resolutions, analysts may choose to down-sample those in a finer resolution, $i$, to match those in a coarser resolution, $i^\prime$. This is a spatial aggregation and if aggregate using the mean, it can be written as -->

<!-- g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}}, -->

<!-- where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid and $n_{i^\prime}$ is the number of observations falls into the coarser grid. Other examples of spatial processing include 1) borrowing information from neighbouring spatial locations to interpolate unobserved locations and 2) fusing variables from ground measures with satellite imageries. -->

@@

`Input`: $\mathbf{x}(s;t)$ where $\mathbf{x} \subseteq \mathbb{R}^p$

`Output`: $\mathbf{y}(s;t)$ where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < p$

`Computation`: Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$ , as the input. It summarises the high-dimensional information into a lower dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < p$, as the output.

The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided in statistical methods. The Standardized Precipitation-Evapotranspiration Index (SPEI) calculates the difference series ($D$) from precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$). This is the only step that differentiates it from the Standardized Precipitation Index (SPI).

When a physical process is absent to describe the phenomenon, multivariate data are often combined using statistical methods, which can be further classified into linear and non-linear techniques. Many social indexes use a linear combination in an additive structure, $y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),$ where $\lambda_i$ is the weight assigned to each individual variable $x_i$. A linear combination can also be preceived as a linear projection of multivariate information through the weight vector. Projecting data from higher to lower dimension unavoidably leads to information loss and the used weights need careful examination to understand their effects on the index and implications for interpretation and decision-making. One can visualise the data projection across different sets of weights through animation using a method called tour.

### 2023-08-15 tidyindex.tex
rework on spatial processing and dimension reduction + proofread

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, frame hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}@{}}
\caption{An notation overview of the input, operation, and output of
each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x(t)\) & \(f[x(.)]\) & \(x^*(t^\prime)\)
where \(t\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x(s)\) & aaaaa & \(x^\#(s^\prime)\) \\
3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) & \(u(s;t)\)
???? \\
4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(v(s;t)\)
where bbbbb \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\) & aaaaa &
\(\mathbf{y}(s;t)\) where \(y \in \mathbb{R}^d\) and \(d < p\) \\
6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
\(F_{\mathbf{\theta}}(.)\) is the cumulated distribution function with
distribution parameter(s) \(\mathbf{\theta}\) & \(p(s,t)\) where
\(p(.)\) denotes the probability value between 0 and 1 \\
7 & Normalising & \(x(s; t)\) & \(\Phi^{-1}[x(s, t)]\) & \(z(s; t)\)
where \(z \sim N(.,.)\) \\
8 & Benchmarking & \(x(s; t)\) & & \(b(s;t)\) \\
9 & Simplification & \(x(s; t)\) & ccccc &
\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

\texttt{Computation}: The temporal processing module takes a single
variable in a set of fixed locations, \(x(t)\), as the input. This step
\emph{transforms} the original series into a new series to incorporate
the temporal neighbouring effect: \(x^*(.) = f[x(.)]\). In temporal
processing, the temporal dimension \(t\) may also transition to
\(t^\prime \subseteq D_{t^\prime}\), allowing for the decomposition of
time into its constituent temporal components. For example, this can be
used to break down daily data into monthly. This gives the output of the
temporal processing module as \(x^*(t^\prime)\).

An example of temporal processing is in Standardized Precipitation Index
(SPI), where monthly precipitation is summed over a rolling window of
size \(k\), also known as the time scale. Assuming an original time
series indexed by \(t= 1, 2, \cdots, T\), the aggregated series, indexed
by \(t^\prime = 1, 2, \cdots, T-k+1\), can be written as
\(x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)\). For SPI,
the choice of the time scale \(k\), is used to control the accumulation
period for the water deficit, enabling the assessment of drought
severity across various types (meteorological, agricultural, and
hydrological).

@@

\texttt{Input}: \(x^l_t(s)\)

\texttt{Output}: \(x^l_t(s^\prime)\) where
\(s^\prime \in D_{s^{\prime}}\)

\texttt{Computation}: Spatial processing may be needed when indexes are
not calculated independently on each collected location or when
variables collected from multiple sources need to be fused before
further processing. The process can be written as a general operation in
the form of

\begin{equation}
x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
\end{equation}

where \(\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}\) is the
associated parameters in the process and \(d_{\theta}\) is the number of
parameter of \(\theta\). An example of spatial processing is to align
variables collected in different resolutions. When variables are
collected at different resolutions, analysts may choose to down-sample
those in a finer resolution, \(i\), to match those in a coarser
resolution, \(i^\prime\). This is a spatial aggregation and if aggregate
using the mean, it can be written as

\begin{equation}
g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}},
\end{equation}

where \(i \in i^\prime\) includes all the cells from the finer
resolution in the coarser grid and \(n_{i^\prime}\) is the number of
observations falls into the coarser grid. Other examples of spatial
processing include 1) borrowing information from neighbouring spatial
locations to interpolate unobserved locations and 2) fusing variables
from ground measures with satellite imageries.

@@

\texttt{Output}: \(x^{l^\prime}(s; t)\), where
\(l^{\prime} = 1, 2, \cdots, L^{\prime}\)

\texttt{Computation}: Dimension reduction summarises the multivariate
information into univariate, which can be denoted as:

\begin{equation}
x_p(\mathbf{s}; \mathbf{t}) \rightarrow x(\mathbf{s}; \mathbf{t}),
\end{equation}

where \(p = 1, 2, \cdots, P\). The combination can be based on
domain-specific knowledge, originated from theories describing the
underlying physical process. For example, the SPEI uses a water balance
model (\(D = P - \text{PET}\)) to calculate the difference series
(\(D\)) from precipitation (\(P\)) and potential evapotranspiration
(\(\text{PET}\)).

Another widely used approach is linear combination, which aggregates a
collection of variables in a linear additive structure, expressed as:
\[x(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),\]

where \(\lambda_p\) denotes the weight assigned to variable \(x_p\).
Most indexes uses an equal weight that sum to 1, which allows each
variable to contribute equally to the index. A linear combination can
also be viewed as a linear projection of multivariate information by the
weight vector. Projecting data from higher to lower dimension inevitably
leads to information loss. For example, two countries can receive
similar index, but their components could significantly differ -- one
with average scores across all components and another with extreme
scores the opposite ends. The selected set of weights need to be
examined to understand its effect on the index and its implications for
decision-making. To do this, analysts can vary the coefficients in the
linear projection to observe how the index value and countries' ranking
change. These changes can also be visualized using a method called tour
by generating an animation of the projections among different sets of
weights.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, enhanced, sharp corners, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
\caption{An notation overview of the input, operation, and output of
each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x(t)\) & \(f_{\Psi}[x(.)]\) &
\(x^*(t^\prime)\) where \(t\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x(s)\) & \(g_{\Theta}[x(.)]\) &
\(x^\#(s^\prime)\) where \(s^\prime \in D_{s^\prime}\) \\
3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) & \(u(s;t)\)
???? \\
4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(v(s;t)\)
where bbbbb \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\) where
\(\mathbf{x} \subseteq \mathbb{R}^p\) & aaaaa & \(\mathbf{y}(s;t)\)
where \(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < p\) \\
6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
\(F_{\mathbf{\theta}}(.)\) is the cumulated distribution function with
distribution parameter(s) \(\mathbf{\theta}\) & \(p(s,t)\) where
\(p(.)\) denotes the probability value between 0 and 1 \\
7 & Normalising & \(x(s; t)\) & \(\Phi^{-1}[x(s, t)]\) & \(z(s; t)\)
where \(z \sim N(.,.)\) \\
8 & Benchmarking & \(x(s; t)\) & & \(b(s;t)\) \\
9 & Simplification & \(x(s; t)\) & ccccc &
\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

\texttt{Computation}: The temporal processing module takes a single
variable in a set of fixed locations, \(x(t)\), as input. This step
transforms the original series into a new series to incorporate the
temporal neighbouring effect: \(x^*(.) = f_{\Psi}[x(.)]\), where
\(\Psi\) is a vector of hyper-parameters associated with the
transformation. In temporal processing, the transition of temporal
dimension from \(t\) to \(t^\prime \subseteq D_{t^\prime}\) allows for
the decomposition of time into its constituent temporal components. For
example, to break down daily data into monthly.

In the Standardized Precipitation Index (SPI), temporal processing is
used to sum the monthly precipitation over a rolling window of size
\(k\), also known as the time scale, prior to the distribution fitting
and normalising module. Assuming an original time series indexed by
\(t= 1, 2, \cdots, T\), the aggregated precipitation series, indexed by
\(t^\prime = 1, 2, \cdots, T-k+1\), can be expressed as
\(x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)\). For SPI,
the choice of the time scale \(k\) is used to control the accumulation
period for the water deficit, enabling the assessment of drought
severity across various types (meteorological, agricultural, and
hydrological).

@@

\texttt{Input}: \(x(s)\)

\texttt{Output}: \(x^\#(s^\prime)\) where
\(s^\prime \in D_{s^{\prime}}\)

\texttt{Computation}: The spatial processing module takes a single
variable with a fixed temporal dimension, \(x(s)\), as input. This step
transforms the variable from the original spatial dimension \(s\) into
the new dimension \(s^\prime \in D_{s^\prime}\) through
\(x^\#(.) = g_{\Theta}(x(.))\), where \(\Theta\) is a vector of
hyper-parameters associated with the transformation. The change of
spatial dimension allows for the alignment of variables collected from
different measurements, such as in-situ stations and satellite imagery,
or originating from different resolutions. This also includes the
aggregation of variables into different levels, such as city, state, and
countries scales.

@@

\texttt{Input}: \(\mathbf{x}(s;t)\) where
\(\mathbf{x} \subseteq \mathbb{R}^p\)

\texttt{Output}: \(\mathbf{y}(s;t)\) where
\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < p\)

\texttt{Computation}: Dimension reduction takes the multivariate
information \(\mathbf{x}(s;t)\), where
\(\mathbf{x} \subseteq \mathbb{R}^p\) , as the input. It summarises the
high-dimensional information into a lower dimension representation
\(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\) and \(d < p\),
as the output.

The transformation can be based on domain-specific knowledge,
originating from theories describing the underlying physical processes,
or guided in statistical methods. The Standardized
Precipitation-Evapotranspiration Index (SPEI) calculates the difference
series (\(D\)) from precipitation (\(P\)) and potential
evapotranspiration (\(\text{PET}\)), using a water balance model
(\(D = P - \text{PET}\)). This is the only step that differentiates it
from the Standardized Precipitation Index (SPI).

When a physical process is absent to describe the phenomenon,
multivariate data are often combined using statistical methods, which
can be further classified into linear and non-linear techniques. Many
social indexes use a linear combination in an additive structure,
\(y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),\) where \(\lambda_i\)
is the weight assigned to each individual variable \(x_i\). A linear
combination can also be preceived as a linear projection of multivariate
information through the weight vector. Projecting data from higher to
lower dimension unavoidably leads to information loss and the used
weights need careful examination to understand their effects on the
index and implications for interpretation and decision-making. One can
visualise the data projection across different sets of weights through
animation using a method called tour.
